DP Problems of Interest

### 2012-10-26 docs/roadmap/draft.tex
Feeding some problems.

\subsubsection*{Todo @TCK}
\begin{verbatim}
- serial implementation for large memory
- optimizations for in-memory computations
- CPU+GPU implementation for S (NS?) problems larger than device memory
\end{verbatim}

>>>>

\subsubsection*{Todo @TCK}
\begin{verbatim}
- continue performance evaluation
- class problem into categories
- think of hashmap implementation. is it useful ?
- CPU+GPU implementation for serial (NS?) problems larger than device memory
- optimizations for in-memory computations

\end{verbatim}

### 2012-10-26 docs/roadmap/problems.tex
Feeding some problems.

% ------------------------------------------------------------------------------------------------
\newpage
\section{Related problems}
The goal of this section is to demonstrate that our framework can accommodate with many problems that we have not considered thoroughly at the design time.

\subsection{Serial problems}
\begin{tabular}{llll} \toprule
\bf Problem & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
Smith-Waterman \footnotesize simple & rectangle & 1 & -- \\
Smith-Waterman \footnotesize affine gap extension & rectangle & 1 & 2 \\
\end{tabular}

\subsection{Non-serial problems}
\begin{tabular}{llll} \toprule
\bf Problem & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
Smith-Waterman \footnotesize arbitrary gap cost & rectangle & 1 & -- \\
Convex polygon triangulation & parallelogram & 1 & -- \\
Matrix chain multiplication & triangle & 1 & -- \\
Nussinov & triangle & 1 & -- \\
Zuker folding & triangle & \color{red} 3? & \color{red} 0? \\
\end{tabular}

{\color{red}
I guess we need between 20-50 DP problems to demonstrate that we identified almost all types of problems. We need to describe common parts between problems (sub categories).

Add additional problems into their related category:
\ul
\item \url{http://en.wikipedia.org/wiki/Dynamic_programming}
\item \url{http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf}
\item \url{http://www.algorithmist.com/index.php/Dynamic_Programming}
\item \url{http://www.cs.uiuc.edu/~jeffe/teaching/algorithms/notes/05-dynprog.pdf}
\item \url{http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf}
\ule

XXX: see comments
}

%
%Dijkstra shortest path, Fibonacci, Balanced 0-1 matrix, Checkerboard, Tower of Hanoi
%Recurrent solutions to lattice models for protein-DNA binding
%Backward induction as a solution method for finite-horizon discrete-time dynamic optimization problems
%Method of undetermined coefficients can be used to solve the Bellman equation in infinite-horizon, discrete-time, discounted, time-invariant dynamic optimization problems
%Many string algorithms including longest common subsequence, longest increasing subsequence, longest common substring, Levenshtein distance (edit distance).
%Many algorithmic problems on graphs can be solved efficiently for graphs of bounded treewidth or bounded clique-width by using dynamic programming on a tree decomposition of the graph.
%The Cocke-Younger-Kasami (CYK) algorithm which determines whether and how a given string can be generated by a given context-free grammar
%Knuth's word wrapping algorithm that minimizes raggedness when word wrapping text
%The use of transposition tables and refutation tables in computer chess
%The Viterbi algorithm (used for hidden Markov models)
%The Earley algorithm (a type of chart parser)
%The Needleman-Wunsch and other algorithms used in bioinformatics, including sequence alignment, structural alignment, RNA structure prediction.
%Floyd's all-pairs shortest path algorithm
%Pseudo-polynomial time algorithms for the subset sum and knapsack and partition problems
%The dynamic time warping algorithm for computing the global distance between two time series
%The Selinger (a.k.a. System R) algorithm for relational database query optimization
%De Boor algorithm for evaluating B-spline curves
%Duckworth-Lewis method for resolving the problem when games of cricket are interrupted
%The Value Iteration method for solving Markov decision processes
%Some graphic image edge following selection methods such as the "magnet" selection tool in Photoshop
%Some methods for solving interval scheduling problems
%Some methods for solving word wrap problems
%Some methods for solving the travelling salesman problem, either exactly (in exponential time) or approximately (e.g. via the bitonic tour)
%Recursive least squares method
%Beat tracking in music information retrieval.
%Adaptive-critic training strategy for artificial neural networks
%Stereo algorithms for solving the correspondence problem used in stereo vision.
%Seam carving (content aware image resizing)
%The Bellman-Ford algorithm for finding the shortest distance in a graph.
%Some approximate solution methods for the linear search problem.
%Kadane's algorithm for the maximum subarray problem.
%	http://en.wikipedia.org/wiki/Dynamic_programming#A_type_of_balanced_0.E2.80.931_matrix
%
%Shortest path in DAGs
%Longest increasing subsequences
%Knapsack
%Shortest path
%All pair shortest paths
%Independent sets in trees
%=> also see exercises for more problems
%	http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf
%
%Longest Increasing Subsequence
%Longest Common Subsequence
%Subset Sum
%Coin Change
%Family Graph
%	http://www.algorithmist.com/index.php/Dynamic_Programming
%
%Longest Increasing Subsequence
%Edit/Levenshtein distance
%Optimal Binary Search Trees
%	http://www.cs.uiuc.edu/~jeffe/teaching/algorithms/notes/05-dynprog.pdf
%
%Independent set on a tree
%0-1 Knapsack
%Floyd-Warshall
%	http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf

@@

We also need to take into account the transfer between main memory (or disk) and device memory. Dependency blocks only need to be read, computed blocks need to be written back. Focusing on the cost blocks only:
\[\begin{array}{rclll}
		b^2 +% computed line writeback (once)
		(b-1)^2 + % diagonal block loads (1 per block)
		\sum\limits_{i=0}^{b-1} i \cdot b % column dependencies loads (for line i)
		&=&\tfrac{1}{2}b^3+\tfrac{3}{2}b^2-2b+1 &\qquad& \rm (Rectangle)
	\\
		\sum\limits_{i=1}^{b} \Big(1+2\cdot(i-1)\Big) \cdot (b+1-i) % on i_th diagonal
		&=& \tfrac{1}{3} b^3 + \tfrac{1}{2}b^2 + \tfrac{1}{6}b &\qquad& \rm (Triangle)
	\\
		\sum\limits_{i=1}^{b} \Big(1+2\cdot(i-1)\Big) \cdot b % on i_th diagonal
		&=& b^3 &\qquad& \rm (Parallelogram)
\end{array}\]

Backtrack blocks need only to be write back and optionally reloaded during backtrack, hence we need to transfer at most:
\[\begin{array}{cccll}
	b^2&+&2b &\qquad& \rm (Rectangle)\\
	\tfrac{1}{2}b^2 + \tfrac{1}{2}b &+& 2b-1 % binary tree with b leafs, hence b-1 interior nodes.
		&\qquad& \rm (Triangle)\\
	b^2 &+& 2b-1 &\qquad& \rm (Parallelogram)

\end{array}\]
Since matrix can grow larger, we store backtrack on 4 bytes. Reusing other assertions from small problems we have $S_C=4, S_B=4, S_I=1, S_W=0$ and $M_D=2^{30}$. From first constraint we have
\[
\frac{n}{b} (1 + 4n + 2\tfrac{n}{b}) \le 2^{29}
\quad\Longleftarrow\quad 5n^2 \le b \cdot 2^{29} \]
Since blocks can not divide cells, we must have $b\le n$ hence $n\le 2^{29}/5 \approx 1.07\cdot 10^8$.

{\color{red} XXX: Continue rewriting here

With $b=8$ we have 337 block transfer, which corresponds to transferring $5.3\times$ the whole matrix, and the problem size can now be (same assertions as for small problems) $2n + 5 \cdot \big(  (n/b)^2 \cdot 2b \big) = 2n+10n^2/8 \le 2^{30} \implies n\le 29307$, so we roughly doubled matrix size at the cost of $12.5\times$ matrix transfer time plus indices overhead.

Assume a different implementation might be worth to support problems order of magnitude larger, hence $n \approx 120K$ thus we need $2^{30} > 2n + 5 \cdot \big(  (n/b)^2 \cdot 2b \big) = $
}

{\color{red} XXX: Do we want to support larger non-serial problem that do not fit in memory? Is it worth implementing ? Argue.}

\subsubsection{Serial problems}

{\color{red}
Split into blocks:\ul
\item Decide the shape of the blocks
\item Decide the size of the blocks
\item Decide of a strategy to store intermediate lines/columns: space/time tradeoff.
\ule

>>>>

% ------------------------------------------------------------------------------------------------
\newpage
\section{Related problems}
The goal of this section is to demonstrate that our framework can accommodate with many problems that we have not considered at the design time.

\subsection{Serial problems}
\begin{tabular}{llcc} \toprule
\bf Problem & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
Smith-Waterman \footnotesize simple & rectangle & 1 & -- \\
Smith-Waterman \footnotesize affine gap extension & rectangle & 3 & -- \\
\href{http://en.wikipedia.org/wiki/Dynamic_programming#Checkerboard}{Checkerboard} & rectangle & 1 & -- \\
\href{http://en.wikipedia.org/wiki/Longest_common_subsequence_problem\#Code_for_the_dynamic_programming_solution}{Longest common subsequence} & rectangle & 1 & -- \\
\href{http://en.wikipedia.org/wiki/Longest_common_substring_problem\#Pseudocode}{Longest common substring} & rectangle & 1 & -- \\
\href{http://en.wikipedia.org/wiki/Levenshtein_distance\#Computing_Levenshtein_distance}{Levenshtein distance} & rectangle & 1 & -- \\
\end{tabular}

\subsection{Non-serial problems}
\begin{tabular}{llcc} \toprule
\bf Problem & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
Smith-Waterman \footnotesize arbitrary gap cost & rectangle & 1 & -- \\
Convex polygon triangulation & parallelogram & 1 & -- \\
Matrix chain multiplication & triangle & 1 & -- \\
Nussinov & triangle & 1 & -- \\
Zuker folding & triangle & \color{red} 3? & \color{red} 0? \\
\href{http://en.wikipedia.org/wiki/CYK_algorithm}{Cocke-Younger-Kasami (CYK)} & triangle & \#rules & -- \\

\end{tabular}

\subsection{Other problems}\ul
\item Dijkstra shortest path: we need a $E\times V$ matrix, along $E$ forall $V$ reduce its distance, problem is serial along $E$, non-serial along $V$ hence of complexity $O(|E|\cdot |V^2|)$ which is far worse than both $O(|V|^2)$ (min-priority queue) and $O(|E|+|V|\log |V|)$ (Fibonacci heap).
\item Fibonacci: this problem is serial 1D. Could be implemented using a placeholder element in one of the matrix dimension.
\item \href{http://archive.ite.journal.informs.org/Vol3No1/Sniedovich/\#dpmodel}{Tower of Hanoi}: 1D non-serial
\item \href{http://www.cs.ust.hk/mjg_lib/bibs/DPSu/DPSu.Files/KnPl81.PDF}{Knuth's word wrapping}: 1D non-serial

\item \href{http://en.wikipedia.org/wiki/Longest_increasing_subsequence#Efficient_algorithms}{Longest increasing subsequence}: serial but binary search algorithm more efficient: $O(n \log n)$.

{\color{red}
\item \href{http://en.wikipedia.org/wiki/Floyd-Warshall_algorithm}{Floyd-Warshall}: <it possible to move the $k$ external loop inside?> Serial with $n$ iterations
\item \href{http://en.wikipedia.org/wiki/Viterbi_algorithm}{Viterbi \footnotesize (hidden Markov models)}: $T$ non-serial iterations over a vector
}
\ule

{\color{red}
Other algorithms to consider/process:\ul
\item \href{http://en.wikipedia.org/wiki/Bellman-Ford_algorithm}{Bellman-Ford} (finding the shortest distance in a graph)
\item \href{http://en.wikipedia.org/wiki/Earley_parser#Pseudocode}{Earley parser} (a type of chart parser)
\item \href{http://en.wikipedia.org/wiki/Needleman-Wunsch_algorithm}{Needleman-Wunsch} (sequences alignement)
\ule
See more in comments
}

Other algorithms used in bioinformatics, including sequence alignment, structural alignment, RNA structure prediction.

% PROBLEM LISTS
% I guess we need between 20-50 DP problems to demonstrate that we identified almost all types of problems.
% http://en.wikipedia.org/wiki/Dynamic_programming
% http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf
% http://www.algorithmist.com/index.php/Dynamic_Programming
% http://www.cs.uiuc.edu/~jeffe/teaching/algorithms/notes/05-dynprog.pdf
% http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf
% ------------------------------------------
% Balanced 0-1 matrix
%Recurrent solutions to lattice models for protein-DNA binding
%Backward induction as a solution method for finite-horizon discrete-time dynamic optimization problems
%Method of undetermined coefficients can be used to solve the Bellman equation in infinite-horizon, discrete-time, discounted, time-invariant dynamic optimization problems
%Many algorithmic problems on graphs can be solved efficiently for graphs of bounded treewidth or bounded clique-width by using dynamic programming on a tree decomposition of the graph.
%The use of transposition tables and refutation tables in computer chess
%Pseudo-polynomial time algorithms for the subset sum and knapsack and partition problems
%The dynamic time warping algorithm for computing the global distance between two time series
%The Selinger (a.k.a. System R) algorithm for relational database query optimization
%De Boor algorithm for evaluating B-spline curves
%Duckworth-Lewis method for resolving the problem when games of cricket are interrupted
%The Value Iteration method for solving Markov decision processes
%Some graphic image edge following selection methods such as the "magnet" selection tool in Photoshop
%Some methods for solving interval scheduling problems
%Some methods for solving word wrap problems
%Some methods for solving the travelling salesman problem, either exactly (in exponential time) or approximately (e.g. via the bitonic tour)
%Recursive least squares method
%Beat tracking in music information retrieval.
%Adaptive-critic training strategy for artificial neural networks
%Stereo algorithms for solving the correspondence problem used in stereo vision.
%Seam carving (content aware image resizing)
%Some approximate solution methods for the linear search problem.
%Kadane's algorithm for the maximum subarray problem.
% =====> http://en.wikipedia.org/wiki/Dynamic_programming#A_type_of_balanced_0.E2.80.931_matrix
%
%Shortest path in DAGs
%Knapsack
%Shortest path
%All pair shortest paths
%Independent sets in trees
%=> also see exercises for more problems
% =====> http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf
%
%Subset Sum
%Coin Change
%Family Graph
% =====> http://www.algorithmist.com/index.php/Dynamic_Programming
%
%Optimal Binary Search Trees
%	http://www.cs.uiuc.edu/~jeffe/teaching/algorithms/notes/05-dynprog.pdf
%
%Independent set on a tree
%0-1 Knapsack
% =====> http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf

@@

We also need to take into account the transfer between main memory (or disk) and device memory. Dependency blocks only need to be read, computed blocks need to be written back. Ignoring the backtrack and focusing only on the cost blocks, the transfers (in blocks) are:
\[\begin{array}{rclll}
		b^2 +% computed line writeback (once)
		(b-1)^2 + % diagonal block loads (1 per block)
		\sum\limits_{i=0}^{b-1} i \cdot b % column dependencies loads (for line i)
		&=&\tfrac{1}{2}b^3+\tfrac{3}{2}b^2-2b+1 &\qquad& \rm (Rectangle)
	\\
		\sum\limits_{i=1}^{b} \Big(1+2\cdot(i-1)\Big) \cdot (b+1-i) % on i_th diagonal
		&=& \tfrac{1}{3} b^3 + \tfrac{1}{2}b^2 + \tfrac{1}{6}b &\qquad& \rm (Triangle)
	\\
		\sum\limits_{i=1}^{b} \Big(1+2\cdot(i-1)\Big) \cdot b % on i_th diagonal
		&=& b^3 &\qquad& \rm (Parallelogram)
\end{array}\]

Putting these two formula together, and using most of the device memory available, we obtain the following results with $S_C=4, S_B\footnote{To deal with larger matrices, backtrack data need to be extended.}=4, S_I=1, S_W=0$ and $M_D=2^{30}$:
\begin{center}\includegraphics[width=14cm]{ns_large.pdf}\end{center}

Given an experimental bandwidth of 5.3743 Gb/s between CPU and GPU, processing matrices one order of magnitude larger (128K) would result in respectively 13\up{(R)}, 8.5\up{(T)} and 25.4\up{(P)} minutes of transfer delay. Extrapolating the preliminary results of small problems, a computation on input of size 128K would require respectively 7 days 13h\up{(R)}, 2 days 22h\up{(T)} and 6 days 10h\up{(P)}, assuming there is no other scalability issues. Although this overhead seems appealing compared to the computation time, the total time blows up (because of the  $O(n^3)$ complexity) and make the processing of such large problem less relevant. Given that real problems -- like RNA folding -- operate at input sizes up to 4096, it would not be of much relevancy to implement a version for larger cases, although perfectly feasible.

\subsubsection{Serial problems}
The serial problem have the interesting property to access to a fixed number of previous elements. These elements

propose a memory consumption scheme

try to setup wavefront size (if needed) => just enlarge the matrix by 1 so that we go wavefront-to-wavefront

{\color{red}

XXX: what's the maximal size of the wavefront ??

XXX: can we avoid to store some matrices and put them in the wavefront ??

Split into blocks:\ul
\item Decide the shape of the blocks
\item Decide the size of the blocks
\item Decide of a strategy to store intermediate lines/columns: space/time tradeoff.
\ule
